Reactive silver inks for antiviral, repellent medical textiles with ultrasonic bleach washing durability compared to silver nanoparticles

### 2022-03-02 Latex/Main.tex
updated paper

There has been a tremendous amount of research on the use of silver nanoparticles as an antimicrobial agent \cite{lansdown:06,Rai:09}
 and their incorporation into 
a range of 
fabrics or masks.\cite{bu_fabrication_2019,perera_morphological_2013, lvm:10, zhong_plasmonic_2020,tremiliosi_ag_2020}
%for healthcare applications
Silver nanoparticles have been shown to inhibit the growth of yeast, \textit{Escherichia coli}, and \textit{Staphylococcus aureus}. \cite{Kim:07mar} %Silicon nanowires decorated with silver nanoparticles have been shown to have antimicrobial effect over 48 hours. \cite{lvm:10} 
%\commentPaul{Some examples with bacteria.}
More recently, silver nanoparticles have been demonstrated to have antiviral properties against enveloped viruses such as severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2) and human immunodeficiency viruses (HIV).\cite{jeremiah_potent_2020,lara_2010} %Silver nanoparticles may interact with viral nucleic acids to disrupt serial viral infection and replication.\cite{jeremiah_potent_2020}
%Silver nanoparticles have been incorporated 
However, nanoparticles on fabrics may have issues with being removed in the wash, resulting in the loss of any associated functionality \cite{Lorenz:12,Impellitteri:09}.
%and other coatings 
%are easily , rendering the surface ineffective , and 
Washed out nanoparticles 
% that is washed out into the environment may 
%The use of silver may %be toxic, which may 
%cause harmful health side effects \cite{Sicherer:13,Stoker:10,Cunha:01} or
may pose a threat to aquatic organisms \cite{Krysanov:10,Ma:13_nanoparticle,Fabrega:11,Pillai:14}
and bioaccumulate in the food chain \cite{Uddin:20}.
Also, the efficacy of silver nanoparticles may be hindered after washing with oxidizing agents such as detergent or bleach.\cite{lansdown:06,Impellitteri:09} 
Silver nanoparticles 
have poor adhesion to the underlying textile fibers due to the small contact area and weak van der Waals forces with the textiles.  Significant research has been dedicated to improving the adhesion forces of silver nanoparticles with textile fibers
such as the use of polymer coatings 
\cite{dastjerdi:2009}, 
polymer binders \cite{Zhang:13feb}, 
binders \cite{Zhou:18}, silanization \cite{shen:2019}, and nanoparticle/polymer composites \cite{Zahran:14}.
%\commentPaul{Do thorough literature search on this.}
Nevertheless, nanoparticle-based methods
are inherently limited by the non-uniformity of the particle deposition and difficulties in controling silver release.\cite{geranio:2009,mitrano:2014}

@@

%Silver ink is a promising alternative for fabric treatments; however, the functionality durability of silver treated fabrics after bleach washing has not been investigated. 
% The use of silver for functional PPE applications must be durable so that 
% functionality is retained after repeated bleach washing.  
This work studies the virucidal activity of
polyethylene terephthalate (PET) knit fabrics coated with a silver amine complex ink \cite{Walker:12}
against herpes simplex viruses (HSV-1). This is the first time in research literature where silver inks have been considered in developing antiviral medical fabrics. We investigate the advantages of silver inks compared to silver nanoparticles for reusable antiviral medical fabrics. 
%Once the silver ink is cured on the fabric, the additional ingredients \commentPaul{Be more specific} \commentsAnthony{These ingredients are proprietary. We could take this sentence out.} of the ink evaporate to leave a pure silver film.   
%The ink is cured thermally
PET fabric material is commonly used for medical and healthcare applications such as gowns, scrubs and caps. \cite{Rigby:97} 
HSV-1 is an enveloped virus about 160 nm in diameter that can cause cold sores and blinding herpetic eye infections.\cite{wald:07} 
The attachment and entry of HSV-1 into cells require the interaction between the viral envelope glycoproteins and cell surface heparan sulfate (HS).\cite{baram-pinto_2009} Silver mimics HS and competes for binding sites of the virus to the cell, inhibiting the virus from cell attachment and entry.\cite{baram-pinto_2009,Galdiero:11}

We show that the reactive silver ink provides better antiviral and bleach wash durability compared to 20 nm diameter silver nanoparticles. Reactive silver ink reduces \textit{in-situ}
on the textile and uniformly covers 
%reduction and uniform silver coverage around 
the microfibers of the fabric.  
This provides for better silver adhesion and uniformity with the fabric microfibers compared to nanoparticles, which adhere to the fabric with weak van der Waals forces. 
%The virus inactivating activity of reactive silver ink is compared with 20 nm silver nanoparticles against HSV-1 in saline.
%Reactive silver ink in saline reduces HSV-1 activity by $1.6 \pm 0.9 \log_{10}$ ($97.3 \pm 5.5 \%$, $p < 0.1$) while 20 nm silver nanoparticles do not show significant activity reduction.  
%Additionally, the virus quantities of PET fabrics coated with silver ink or silver nanoparticles is compared. 
PET fabrics coated with 20 nm silver nanoparticles %lose majority of its 
show a $1.1 \pm 0.6$ log$_{10}$ ($92.4 %\pm 2.5 
\%$, $p < 0.05$) log reduction in virus quantities compared to control fabric, 
while PET fabrics coated with reactive silver inks reduce virus quantities by $2.3 \pm 0.8$ log$_{10}$ ($99.5 %\pm 3.0 
\%$, $p < 0.001$). 
%The reduction of virus quantities on the fabric is likely from the liquid repellent properties, since both treatments render the fabric highly hydrophobic. The liquid containing the virus does not penetrate the treated fabric and a reduction of virus quantities is observed compared to controls.

However, we observe the performance of both silver fabric treatments can be significantly hindered after extended periods of ultrasonic bleach washing. 
%\commentPaul{Any ideas on the mechanism why reactive silver ink would be able to inactivate but silver nanoparticles not.}
%\commentPaul{Give percentage.}
%\commentPaul{Give $p$ number from t-test?}
%$2.3 \pm 0.8$ log, $2.7 \pm 0.8$ log, $2.7 \pm 0.8$ log and $2.8 \pm 0.7$ log PFU per mL, respectively.
%for enhanced stability. 
% In this work, we compare the anti-viral properties of a reactive silver ink in phosphate buffered saline (PBS) and on polyester (PET) fabric with silver nanoparticles. 
An additional polydimethylsiloxane (PDMS) thin film is used protect the silver layer and increase the liquid repellent properties of the fabric to further improve %decontamination
the bleach wash durability of the antiviral functionality. 
%The antiviral property of silver-PDMS fabric treatments after washing with bleach has not been tested previously. 
PET fabric samples coated with reactive silver ink and PDMS reduce virus quantities by $1.7 \pm 0.2$ log$_{10}$ ($98.2 %\pm 0.8 
\%$, $p < 0.01$) after 300 minutes of ultrasonic bleach washing. We demonstrate the combination of reactive silver ink with PDMS as a PET fabric treatment that adds durable, superhydrophobic and antiviral properties for reusable PPE applications.

@@

\begin{figure}[H]
       \centering
    \includegraphics[width= \linewidth]{Figures/fig1_char.pdf}
   % \vspace{.1in}
   \caption[Sample characterization of silver coated fabric]{a) Schematic of silver treatment on PET fabric (b-d) SEM images of b) untreated PET c) PET-NP and d) PET-Ink. e-f) XRD spectra of e) PET-NP and f) PET-Ink samples.}
   %\commentPaul[inline]{Perhaps (c) and (d) should be (c)(i) and (ii).}
%    \commentPaul[inline]{Change text in (a) to ``Coat Ag Nanoparticles or Ink".}
    % \commentPaul[inline]{Add ``Ag Ink or Nanoparticle-Coated PET" to schematic.  Add another arrow (maybe two rows) with final sample.}
   %\commentPaul[inline]{Change text in (a) to ``Ag Nanoparticles or Ink".}
   %\commentPaul[inline]{Change (f) order to Control, NP, ink.  Make line in (i) and (ii) same blue color.  Change order of legend in (e) to PET and then with -NP and with -Ink.}
%   \commentPaul[inline]{Change line format so not overlapping so much.}

@@

\begin{figure}[H]
       \centering
    \includegraphics[width= \linewidth]{Figures/fig_wash1.pdf}
   % \vspace{.1in}
\caption[Wash Durability of silver coated fabric]{
a) Wetting characterization for PET, PET-NP, and PET-Ink samples. b) Silver leaching analysis as a function of bleach washing time c) Optical images of (i) PET-NP and (ii) PET-Ink samples as a function of bleach washing time. d) Virus inhibition assay results of silver treatments in solution.} %FTIR spectra of etched PET GO-ODA + PDMS before and after 33 washes. e-f) Optical images of water (blue) and human saliva (yellow) droplets on i) EPGO and ii) EPGOP after (e) 250 abrasion cycles or (f) 33 bleach washing cycles.  %}
%\commentPaul[inline]{Perhaps change axes in (b) so max is 100 and min is 0.}
 %\commentsAnthony[inline]{i think it looks better with some space}
 %\commentPaul[inline]{The issue is that 0 and 100 are also the theoretical max and min.  I think you can change the plot so that the datapoint and error bar is on top of the axis if that is why you think it doesn't look good.}
 %\commentPaul[inline]{Maybe merge Table in (a) with Figure 1? (a) doesn't seem connected to rest of Figure.}
 %\commentPaul[inline]{Move 1(f) into this Figure as (d) to match order below.}

\label{fig:durable}
\end{figure}

>>>>

There has been a tremendous amount of research on the use of silver nanoparticles as an antimicrobial agent \cite{lansdown:06,Rai:09}
 and their incorporation into 
a range of 
fabrics or masks.\cite{bu_fabrication_2019,perera_morphological_2013, lvm:10, zhong_plasmonic_2020,tremiliosi_ag_2020,abulikemu:2022} 
%for healthcare applications
Silver nanoparticles have been shown to inhibit the growth of yeast, \textit{Escherichia coli}, and \textit{Staphylococcus aureus}. \cite{Kim:07mar} %Silicon nanowires decorated with silver nanoparticles have been shown to have antimicrobial effect over 48 hours. \cite{lvm:10} 
%\commentPaul{Some examples with bacteria.}
More recently, silver nanoparticles have been demonstrated to have antiviral properties against enveloped viruses such as severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2) and human immunodeficiency viruses (HIV).\cite{jeremiah_potent_2020,lara_2010,elechiguerra_interaction_2005} %Silver nanoparticles may interact with viral nucleic acids to disrupt serial viral infection and replication.\cite{jeremiah_potent_2020}
%Silver nanoparticles have been incorporated

However, nanoparticles on fabrics may have issues with being removed in the wash, resulting in the loss of any associated functionality \cite{Lorenz:12,Impellitteri:09}.Recently published articles fail to investigate the washing durability of silver nanoparticles incorporated into fabrics for the use of personal protective equipment.\cite{abulikemu:2022,zhong_plasmonic_2020}
%and other coatings 
%are easily , rendering the surface ineffective , and 
Washed out nanoparticles 
% that is washed out into the environment may 
%The use of silver may %be toxic, which may 
%cause harmful health side effects \cite{Sicherer:13,Stoker:10,Cunha:01} or
may pose a threat to aquatic organisms \cite{Krysanov:10,Ma:13_nanoparticle,Fabrega:11,Pillai:14}
and bioaccumulate in the food chain \cite{Uddin:20}.
Also, the efficacy of silver nanoparticles may be hindered after washing with oxidizing agents such as detergent or bleach.\cite{lansdown:06,Impellitteri:09} 
Silver nanoparticles 
have poor adhesion to the underlying textile fibers due to the small contact area and weak van der Waals forces with the textiles.  Significant research has been dedicated to improving the adhesion forces of silver nanoparticles with textile fibers
such as the use of polymer coatings 
\cite{dastjerdi:2009}, 
polymer binders \cite{Zhang:13feb}, 
binders \cite{Zhou:18,abulikemu:2022}, silanization \cite{shen:2019}, and nanoparticle/polymer composites \cite{Zahran:14}.
%\commentPaul{Do thorough literature search on this.}
Nevertheless, nanoparticle-based methods
are inherently limited by the non-uniformity of the particle deposition and difficulties in controling silver release.\cite{geranio:2009,mitrano:2014}

@@

%Silver ink is a promising alternative for fabric treatments; however, the functionality durability of silver treated fabrics after bleach washing has not been investigated. 
% The use of silver for functional PPE applications must be durable so that 
% functionality is retained after repeated bleach washing.  
This work studies the virucidal activity of
polyethylene terephthalate (PET) knit fabrics coated with a silver amine complex ink \cite{Walker:12}
against herpes simplex viruses (HSV-1). This is the first time in research literature where silver inks have been considered in developing antiviral medical fabrics. We investigate and discuss the advantages of silver inks compared to silver nanoparticles for reusable antiviral medical fabrics. 
%Once the silver ink is cured on the fabric, the additional ingredients \commentPaul{Be more specific} \commentsAnthony{These ingredients are proprietary. We could take this sentence out.} of the ink evaporate to leave a pure silver film.   
%The ink is cured thermally
PET fabric material is commonly used for medical and healthcare applications such as gowns, scrubs and caps. \cite{Rigby:97} 
HSV-1 is an enveloped virus about 160 nm in diameter that can cause cold sores and blinding herpetic eye infections.\cite{wald:07} 
The attachment and entry of HSV-1 into cells require the interaction between the viral envelope glycoproteins and cell surface heparan sulfate (HS).\cite{baram-pinto_2009} Silver mimics HS and competes for binding sites of the virus to the cell, inhibiting the virus from cell attachment and entry.\cite{baram-pinto_2009,Galdiero:11}

We show that the reactive silver ink provides better antiviral and bleach wash durability compared to 20 nm diameter silver nanoparticles. The size of the silver nanoparticles is considered as it is the maximum size to offer virus inhibition with HIV-1. \cite{elechiguerra_interaction_2005} Silver nanoparticles of similar size on masks are shown to inhibit human coronavirus \cite{abulikemu:2022}. The reactive silver ink reduces \textit{in-situ}
on the textile and uniformly covers 
%reduction and uniform silver coverage around 
the microfibers of the fabric.  
This provides for better silver adhesion and uniformity with the fabric microfibers compared to nanoparticles, which adhere to the fabric with weak van der Waals forces. 
%The virus inactivating activity of reactive silver ink is compared with 20 nm silver nanoparticles against HSV-1 in saline.
%Reactive silver ink in saline reduces HSV-1 activity by $1.6 \pm 0.9 \log_{10}$ ($97.3 \pm 5.5 \%$, $p < 0.1$) while 20 nm silver nanoparticles do not show significant activity reduction.  
%Additionally, the virus quantities of PET fabrics coated with silver ink or silver nanoparticles is compared. 
PET fabrics coated with 20 nm silver nanoparticles %lose majority of its 
show a $1.1 \pm 0.6$ log$_{10}$ ($92.4 %\pm 2.5 
\%$, $p < 0.05$) log reduction in virus quantities compared to control fabric, 
while PET fabrics coated with reactive silver inks reduce virus quantities by $2.3 \pm 0.8$ log$_{10}$ ($99.5 %\pm 3.0 
\%$, $p < 0.001$). 
%The reduction of virus quantities on the fabric is likely from the liquid repellent properties, since both treatments render the fabric highly hydrophobic. The liquid containing the virus does not penetrate the treated fabric and a reduction of virus quantities is observed compared to controls.

However, we observe the performance of both silver fabric treatments can be significantly hindered after extended periods of ultrasonic bleach washing. 
%\commentPaul{Any ideas on the mechanism why reactive silver ink would be able to inactivate but silver nanoparticles not.}
%\commentPaul{Give percentage.}
%\commentPaul{Give $p$ number from t-test?}
%$2.3 \pm 0.8$ log, $2.7 \pm 0.8$ log, $2.7 \pm 0.8$ log and $2.8 \pm 0.7$ log PFU per mL, respectively.
%for enhanced stability. 
% In this work, we compare the anti-viral properties of a reactive silver ink in phosphate buffered saline (PBS) and on polyester (PET) fabric with silver nanoparticles. 
An additional polydimethylsiloxane (PDMS) thin film is used protect the silver layer and increase the liquid repellent properties of the fabric to further improve %decontamination
the bleach wash durability of the antiviral functionality. 
%The antiviral property of silver-PDMS fabric treatments after washing with bleach has not been tested previously.
When the silver layer is protected by the polymer, the silver ink treated fabric retains 42\% more silver than silver nanoparticle treated fabrics after bleach washing.
PET fabric samples coated with reactive silver ink and PDMS reduce virus quantities by $1.7 \pm 0.2$ log$_{10}$ ($98.2 %\pm 0.8 
\%$, $p < 0.01$) after 300 minutes of ultrasonic bleach washing. 
We demonstrate the combination of reactive silver ink with PDMS as a PET fabric treatment that adds durable, superhydrophobic and antiviral properties for reusable PPE applications.

@@

\begin{figure}[H]
       \centering
    \includegraphics[width= \linewidth]{Figures/fig1_char.pdf}
   % \vspace{.1in}
   \caption[Sample characterization of silver coated fabric]{a) Schematic of silver treatment on PET fabric (b-d) SEM images of b) untreated PET c) PET-NP and d) PET-Ink. e-f) XRD spectra of e) PET-NP and f) PET-Ink samples.}

@@

\begin{figure}[H]
       \centering
    \includegraphics[width= \linewidth]{Figures/fig_wash1.pdf}
   % \vspace{.1in}
\caption[Wash Durability of silver coated fabric]{
a) Wetting characterization for PET, PET-NP, and PET-Ink samples. b) Silver leaching analysis as a function of bleach washing time c) Optical images of (i) PET-NP and (ii) PET-Ink samples as a function of bleach washing time. d) Virus inhibition assay results of silver treatments in solution.} 
\label{fig:durable}
\end{figure}
